Dynamic capabilities and competitive advantage

### 2019-11-15 article4_jbr.rnw
WOrking on post hoc analysis

<<load, results="asis", echo=FALSE >>=

setwd("/Users/larshovdanmolden/Documents/git/article4")
library(xtable)
#load("./data/resart4.RData")
r <- main$resmat
rm  <- main$resmatmed
l  <- main$loadings

foolars=0.99

@
%% Extant literature on dynamic capabilities is battling with a fundamental conception and
%% prediction of the theory, namely if, when and how dynamic capabilities can lead to competitive
%% advantage.
%% This paper is about how apsobritve capacity can come about as strategic posture
%https://arrow.dit.ie/cgi/viewcontent.cgi?article=1009&context=buschmancon

% paper about the DC for MNC in india an dimpact on CA
%https://journals.sagepub.com/doi/abs/10.1177/0971890717701781

Extant literature on dynamic capabilities are increasingly focusing on how firm resources
are changed by dynamic capabilities as a response to changes in the environment
\citep{Schilke2018,Eriksson2014}, thus placing firm resources as a mediator between
dynamic capabilities and competetive advantage \citep{Helfat2007}. Less focus has been
placed on the nature of these underlying firm resources (ref) and how they are capturing a
rather heterogeneous bulk of routines, capabilities, assets and capital. Hence, there is a
need to disentangle the black box of firm resources to provide a more nuanced
understanding of different types of firm resources and how they interact to shape firm
competitive advantage \cite{Schilke2018}. Indeed, this gap in the literature is explicitly
emphasized by \cite{Schilke2018} where they 'see an interesting opportunity for future
work to add greater richness to our understanding of the mechanism of resource-base
change, given this mechanism’s central role in many foundational works (..) and the
diverse ways in which resource changes can potentially come about' (p. 419).

@@

However, as \cite{Verona2011} points out: 'This approach misses the fundamental aspect
that organizational capabilities specific to managing change can hardly be reduced to the
management of behavioral change' (ibid p. 538). This opens for \emph{non-behavioral
  objects of change} which are defined as 'behavioral antecedents, such as cognition
(e.g. cognitive frames) and motivation' (ibid p. 539). This concept entails the 'how
decisions or actions are shaped, subverted, and/or transformed by emotions; and it
concerns how emotions become embedded in cultural and political practices that determine
the "way we do things here"' (Vince and Gabriel 2011 . ch 15). One particular dimension of
these non-behavioral objects of change particularly conducive to the context of
generating strategic change is the \emph{strategic cognition} of the
organization. Strategic cognition can be defined as conceptual structures in the minds of
the individuals that encapsulate a shared understanding of the reality the
individuals face with respect to the strategic goal of the organization \footnote{This
  definition is based on the definition of \cite{Lakoff1987} definition of the more generic
  'cognitive representation' and is adapted to fit the pursuit of strategic goals such as
  competetive advantage which is the aim of this paper}. This strategic
cognition is the result of a mental process that underlie internal
audicences adoption of new strategic representation and identity code to move toward a
new or existing goal \citep{Gavetti2012}. Although strategic cognition may lead to
behavioral outcome, the nature of the concept itself is tied to cognition rather than
actual latent action \citep{Zollo2016}. Thus, dynamic capabilities are not sufficiently
creating strategic change through its role as 'conductor'. Additionally they need to
exercise a role of 'coach' to adapt the strategic cognition of the organization to create
a shared understanding of the sense and necessity of the strategic goals of the
orientation.

@@

There are no established measurement models of dynamic capabilities
\citep{McKelvie2009,Schilke2018}.Therefore this study builds on a
mixture of qualitative case study methodology, literature review and
statistical techniques to develop and refine measures of DCs. First,
exploratory qualitative interviews, using a semistructured interview
guide, were conducted with management representatives from 10
R\&D/innovative firms. The aim of the interviews was to get an
overview of each firm's innovation and development processes, in
particular the processes related to its dynamic capabilities and the
management of its resources. Themes raised in the interviews were
about network, co-operation with external R\&D-institutions, learning
in the firm, adaptation and changes in the firm. We interviewed SMEs
and larger firms, and the industries varied from high-tech and ICT to
publishing.  Based on the interviews and an extensive literature
review, statements identified as descriptions of dynamic capabilities
and resources were developed and included in a questionnaire.

@@

Respondents were asked to indicate to what extent each statement fitted a description of
their business. We adopted one-sided seven point Likert scale where: 1 = strongly disagree
and 7 = strongly agree. We built on prior studies where items measuring OCs have been
measured relative to competitors \citep{McKelvie2009}.

@@

@@

To test our hypotheses we then built a path-model from the latent
variables we constructed. This resulted in four models. The results
from the structural model is reported in figure \ref{fig:res} and
table \ref{tab:res}. To estimate the standard errors we ran
simulations using bootstrapping and reported the simulation number,
standard errors and corresponding p-values. To evaluate the models we
compared $R^2$ and $Q^2$ statistics, as well as changes in Baysian
Information Criteria (BIC) for model comparison. All these are
reported in table \ref{tab:res}. Finally, we tested the model for
nonlinear effets, endogeneity and unbserved heterogeneity using
Ramsey, Gaussian Copula approach and Information Criteria test
respectively. These results are available in the appendix.

@@

\section{Results}

\begin{table}[t]\label{tab:cor}
  \caption{Descriptive statistics and correlation between key constructs and control
  variables.}
  <<corchild, child='./figure/cortab.rnw',results="asis">>=
@

\begin{table}[t]\label{tab:res}
  \caption{PLS-SEM results for key constructs}

  \scriptsize
  <<child-demo, child='./figure/restab.rnw'>>=
@
   \begin{tablenotes}
      \tiny
      \item Standard-errors estimated using bootstrapping with 10 000 simulations.
      \item Firm level control variables include firm size, firm age,
        profit margin, firm assets.

    \end{tablenotes}
  \end{table}

 <<pathchild, child='./figure/medanalysisall.rnw',results="asis">>=
@

>>>>

<<loading, results="asis", echo=FALSE >>=
setwd("/Users/larshovdanmolden/Documents/git/article4")
library(xtable)
load("./data/resart4.RData")
r <- main$resmat
rm  <- main$resmatmed
l  <- main$loadings
l <- main$loadings
#foolars=0.99
@

%% Extant literature on dynamic capabilities is battling with a fundamental conception and
%% prediction of the theory, namely if, when and how dynamic capabilities can lead to competitive
%% advantage.
%% This paper is about how apsobritve capacity can come about as strategic posture
%https://arrow.dit.ie/cgi/viewcontent.cgi?article=1009&context=buschmancon

% paper about the DC for MNC in india an dimpact on CA
%https://journals.sagepub.com/doi/abs/10.1177/0971890717701781
BLA
Extant literature on dynamic capabilities are increasingly focusing on how firm resources
are changed by dynamic capabilities as a response to changes in the environment
\citep{Schilke2018,Eriksson2014}, thus placing firm resources as a mediator between
dynamic capabilities and competetive advantage \citep{Helfat2007}. Less focus has been
placed on the nature of these underlying firm resources (ref) and how they are capturing a
rather heterogeneous bulk of routines, capabilities, assets and capital. Hence, there is a
need to disentangle the black box of firm resources to provide a more nuanced
understanding of different types of firm resources and how they interact to shape firm
competitive advantage \cite{Schilke2018}. Indeed, this gap in the literature is explicitly
emphasized by \cite{Schilke2018} where they 'see an interesting opportunity for future
work to add greater richness to our understanding of the mechanism of resource-base
change, given this mechanism’s central role in many foundational works (..) and the
diverse ways in which resource changes can potentially come about' (p. 419).

@@

However, as \cite{Verona2011} points out: 'This approach misses the fundamental aspect
that organizational capabilities specific to managing change can hardly be reduced to the
management of behavioral change' (ibid p. 538). This opens for \emph{non-behavioral
  objects of change} which are defined as 'behavioral antecedents, such as cognition
(e.g. cognitive frames) and motivation' (ibid p. 539). This concept entails the 'how
decisions or actions are shaped, subverted, and/or transformed by emotions; and it
concerns how emotions become embedded in cultural and political practices that determine
the "way we do things here"' (Vince and Gabriel 2011 . ch 15). One particular dimension of
these non-behavioral objects of change particularly conducive to the context of
generating strategic change is the \emph{strategic cognition} of the
organization. Strategic cognition can be defined as conceptual structures in the minds of
the individuals that encapsulate a shared understanding of the reality the
individuals face with respect to the strategic goal of the organization \footnote{This
  definition is based on the definition of \cite{Lakoff1987} definition of the more generic
  'cognitive representation' and is adapted to fit the pursuit of strategic goals such as
  competetive advantage which is the aim of this paper}. This strategic
cognition is the result of a mental process that underlie internal
audiences adoption of new strategic representation and identity code to move toward a
new or existing goal \citep{Gavetti2012}. Although strategic cognition may lead to
behavioral outcome, the nature of the concept itself is tied to cognition rather than
actual latent action \citep{Zollo2016}. Thus, dynamic capabilities are not sufficiently
creating strategic change through its role as 'conductor'. Additionally they need to
exercise a role of 'coach' to adapt the strategic cognition of the organization to create
a shared understanding of the sense and necessity of the strategic goals of the
orientation.

@@

There are no established measurement models of dynamic capabilities
\citep{McKelvie2009,Schilke2018}.Therefore this study builds on a mixture of qualitative
case study methodology, literature review and statistical techniques to develop and refine
measures of DCs. First, exploratory qualitative interviews, using a semi-structured
interview guide, were conducted with management representatives from 10 R\&D/innovative
firms. The aim of the interviews was to get an overview of each firm's innovation and
development processes, in particular the processes related to its dynamic capabilities and
the management of its resources. Themes raised in the interviews were about network,
co-operation with external R\&D-institutions, learning in the firm, adaptation and changes
in the firm. We interviewed SMEs and larger firms, and the industries varied from
high-tech and ICT to publishing.  Based on the interviews and an extensive literature
review, statements identified as descriptions of dynamic capabilities and resources were
developed and included in a questionnaire.

@@

Respondents were asked to indicate to what extent each statement fitted a description of
their business. We adopted one-sided seven point Likert scale where: 1 = strongly disagree
and 7 = strongly agree. We built on prior studies where items measuring operating
capabilities have been measured relative to competitors \citep{McKelvie2009}.

The nine year lag between our points of measurement has its distinct advantages with
respect to capturing long term trajectory of the firms. For example, the operating
capabilities at time $t=1$ is very likely a function of operating capabilities at time
$t=0$ as well as other factors such as dynamic capabilities. Indeed, extant literature
argues for this path dependency of resources and routines
\citep{Helfat2007,Winter2003}. These capabilities are said to be serial-correlated and
path dependent. However, as dynamic capabilities and operating capabilities are distinctly
different concepts, this kind of path dependency is also not equally clear in this
regard. We could have firms with very high levels of dynamic capabilities at $t=0$ and
equally a poorer level at time $t=1$. Because of the long time lag a lot of factors could
have influenced this decline, and indeed a rather high number of over firms exhibit this
characteristic of declining dynamic capabilities over time. As we regard dynamic
capabilities as 'the first derivative' \citep{Winter2003} (i.e. the rate of change of the
operating capabilites) a large initial value (at time $t=0$) could over estimate the
expected change of the underlying capability at time $t=1$. Moreover, based on the scale
applied (Likert from 1 through 7) a high initial value leaves nowhere to move but
downwards. This is not to say that those firms are likely to develop poorer operating
routines. Nor is to say that those with low dynamic capabilities at time $t=0$ are
rendered useless in developing capabilities going forward towards $t=1$. Rather it
suggests that a long time span may include unnecessary noise in the model. Thus, we
estimated an average of the dynamic capabilities of the firm between time $t=0$ and $t=1$
as a proxy for a more representative measure of the rate of change. Hence, our construct
of $DC_0$ is actually a measure between times $t=0$ and $t=1$. As a robustness check we
repeated the full analysis using 2005 measure as our $DC_0$. This did not severely alter
the conclusions we draw in this paper, but had some minor effects on the significance of
the path between $DC_0$ and $COG_1$. The full alternative analysis is included in the
appendix.

@@

All the constructs and the control variables are presented with descriptive statistics and
pairwise correlations. The main estimation was tested for multicolinearity using the
variance inflation test (VIF) with no severe multicolinearity detected. This is also
suggested by the correlation matrix.

\begin{table}[t]
  \caption{Descriptive statistics and correlation between key constructs and control
  variables.}
\label{tab:cor}
  <<corchild, child='./figure/cortab.rnw',results="asis">>=
@

@@

To test our hypotheses we then built a path-model from the latent variables we
constructed. This resulted in four models. The results from the structural model is
reported in figure \ref{fig:res} and table \ref{tab:res}. To estimate the standard errors
we ran simulations using bootstrapping and reported the simulation number, standard errors
and corresponding p-values. To evaluate the models we compared $R^2$ and $Q^2$ statistics,
as well as changes in Baysian Information Criteria (BIC) for model comparison. All these
are reported in table \ref{tab:res}. Finally, we tested the model for nonlinear effects,
endogeneity and unbserved heterogeneity using Regression Equation Specification Error Test
(RESET) , Gaussian Copula approach and Information Criteria test respectively. These
results are available in the appendix.

@@

\section{Results}
The main results of our analysis are depicted both in the graphical path diagrams (figure
\ref{fig:res}) and the regression table in table \ref{tab:res}. The empirical models are
presented from left to right starting with the direct model. The main effect from $DC_0$
to $CA_1$, controlled for $CA_0$, firm size, age, profit margin and assets, is positive
and significant at the onset. Increasingly, by adding mediating variables as suggested by
the theory, this effect diminishes to an extent suggesting that partial or full mediation
is taking place. The effect decreases from BLA to BLA and turns insignificant. The
mediating variables $CAP_1$ and $COG_1$ are both positive and significantly related to
$CA_1$. This suggests that the effect between dynamic capabilities and competitive
advantage is mediated by these two factors separately, as well as together. These findings
both support H1 and H2. Moreover, the model becomes increasingly better both in terms if
explanatory power ($R^2$), predictive power ($Q^2$) and the Bayesian Information Criteria
(BIC).

\begin{table}[t]
  \caption{PLS-SEM results for key constructs}
\label{tab:res}
  \scriptsize
  <<child-demo, child='./figure/restab.rnw'>>=
@
   \begin{tablenotes}
      \tiny
      \item Standard-errors estimated using bootstrapping with 10 000 simulations.
      \item Firm level control variables include firm size, firm age,
        profit margin, firm assets.
    \end{tablenotes}
  \end{table}

One striking finding, however, is that the improvement in the model between
model 3 and 4 is neglectable, whereas it represents a considerable improvement between 2 and 3. This
suggests that strategic cognition represents a stronger mediator than operational
capabilities. However, the difference is too small to render further analysis all but
conjecture on our part. Thus, we wanted found it necessary to explore the joint mediation
between $CAP_1$ and $COG_1$.

\subsection{Post-hoc analysis: joint mediation}

From standard production theory we know that the isoquants
between two factors of input rarely are angled (Leontief isoquants) suggesting that the
factors are total complements to eachother. Rather there are trade offs between these
factors suggesting a non-linear relationship. One way to investigate this is by means of
joint mediation where (USE MODERATED MEDIATION RATHER). 
 <<pathchild, child='./figure/medanalysisall.rnw',results="asis">>=
@
